\documentclass[UKenglish]{article}

%\usepackage[utf8]{inputenc}
\RequirePackage[no-math]{fontspec}

\usepackage[small]{dgruyter}
\usepackage{microtype}
\usepackage{enumerate}
\usepackage{eqnarray}
\usepackage{colortbl}
\usepackage{soul}

% \usepackage[firstpage]{draftwatermark}
% \SetWatermarkFontSize{30pt}
% \SetWatermarkHorCenter{300pt}
% \SetWatermarkColor[rgb]{0.9,0.5,0.5}
% \SetWatermarkText{\textsf{Submitted April 2017. Do not quote.}}

\newcommand{\eg}{e.\,g.,}
\newcommand{\Eg}{E.\,g.,}
\newcommand{\ie}{i.\,e.,}
\newcommand{\Ie}{I.\,e.,}
\newcommand{\Sub}[1]{\ensuremath{\mathrm{_{#1}}}}
\newcommand{\Sup}[1]{\ensuremath{\mathrm{^{#1}}}}
\newcommand{\Subsf}[1]{\ensuremath{\mathsf{_{#1}}}}
\newcommand{\Supsf}[1]{\ensuremath{\mathsf{^{#1}}}}
\newcommand{\pPB}{p\Subsf{PB}}
\newcommand{\mpPB}{\ensuremath{p_{\text{PB}}}}

\newcommand{\NACb}{NAC\Sub{bare}}
\newcommand{\NACa}{NAC\Sub{adj}}
\newcommand{\PGCd}{PGC\Sub{det}}
\newcommand{\PGCa}{PGC\Sub{adj}}

\newcommand*\Rot{\rotatebox{75}}

\definecolor{lg}{rgb}{.8,.8,.8}
\newcommand{\Dim}{\cellcolor{lg}}

\usepackage{gb4e-}

\usepackage{bbding}
\newcommand{\CheckIt}{\CheckmarkBold}
\newcommand{\CrossIt}{\XSolidBold}


\usepackage{natbib}
\bibliographystyle{./unified}

\usepackage{lineno}

\begin{document}

% \runninglinenumbers
% \modulolinenumbers[1]
% \setlength\linenumbersep{20pt}

  %\articletype{...}

  %\author*[1]{Roland Schäfer}
  %\runningauthor{Roland Schäfer. Submitted April 2017. Do not quote.}
  %\affil[1]{Freie Universität Berlin}

  \title{Abstractions and exemplars: the measure noun phrase alternation in German}
  \runningtitle{Abstraction and exemplars: the measure noun phrase alternation German}
  %\subtitle{...}

  \abstract{In this paper, an alternation in German measure noun phrases is examined under a varying-abstraction perspective.
    In a specific measure NP construction, the embedded kind-denoting noun either agrees in case with the measure noun (\textit{eine Tasse guter Kaffee} `a cup of good coffee') or it stands in the genitive (\textit{eine Tasse guten Kaffees}).
    Each of the two alternants is syntactically similar to a non-alternating variant.
    I propose a prototype model which assigns a common prototypical meaning to each of the alternants and its corresponding non-alternating construction.
    Based on that, I predict that lexical, morpho-syntactic, and stylistic features help to predict the choice of the alternant.
    A large corpus study is presented which supports this theory.
    However, in addition to the prototype effects, an exemplar effect is also shown to influence the choice, namely the relative frequencies with which lemmas occur in the non-alternating constructions.
    I argue that allowing both prototype and exemplar effects is more adequate than following radical prototype or exemplar approaches.
    It is also verified in two experiments that the corpus-derived model corresponds to the behaviour of native speakers.
    The weak effect size of the experimental validation is discussed in the context of corpus-based cognitive linguistics and the validation of corpus-derived models.
  }
  \keywords{prototypes vs.\ exemplars, corpus methods and experimental validation, alternations, hierarchical models, pseudo-partitives, German}
  %\classification[PACS]{...}
  %\communicated{...}
  %\dedication{...}
  %\received{...}
  %\accepted{...}
  %\journalname{Submitted April 2017, revised version submitted January 2018. Do not quote.}
  %\journalyear{}
  %\journalvolume{...}
  %\journalissue{...}
  %\startpage{...}
  %\aop
  %\DOI{...}

  
\maketitle


\include{sections/1intro}
\include{sections/2descriptive}
\include{sections/3corpus}
\include{sections/4experiments}
\include{sections/5conclusions}



% \begin{acknowledgement}
%   I thank (in alphabetical order) Felix Bildhauer, Susanne Flach, Elizabeth Pankratz, Samuel Reichert, Ulrike Sayatz, and Christian Zimmer for valuable discussions and comments.
%   Also, I would like to thank the reviewers for Cognitive Linguistics as well as associate editor Dagmar Divjak for insightful comments which helped to improve the quality of this paper significantly.
%   Furthermore, I thank Ulrike Sayatz for helping me to recruit the participants for the experiments.
%   Elizabeth Pankratz thankfully also fixed my English.
%   Finally, I am grateful to my student assistants Kim Maser for her work on the annotation of the concordances and Luise Rißmann for supervising most of the experiments.
%   The research presented here was made possible in part through funding from the \textit{Deutsche Forschungsgemeinschaft} (DFG, personal grant SCHA1916/1-1).
% \end{acknowledgement}


\bibliography{rs}

\end{document}

